Notes on the low density parameters in {\tt MAESTRO}

### 2011-06-23 docs/initial_models/initial_models.tex
use maestrosymbols

Here we briefly describe the various routines for generating initial models
for the main MAESTRO problems and how the initial model is used to initialize
both the base state and the full Cartesian state.

@@

We have found that for the best performance, the MAESTRO initialization procedure should be given model data with the same resolution as the base state resolution, $\Delta r$.  For planar problems, $\Delta r = \Delta x$.  For multilevel planar problems, we use $\Delta r$ equal to $\Delta x$ at the finest resolution.  For spherical problems we set $\Delta r = \Delta x/\mathtt{drdxfac}$.  We generate our initial model either analytically or from raw data,
 $\rho^{\raw}, T^{\raw}, p^{\raw}$, and $X^{\raw}$.  
Here is the raw data file for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt test\_convect} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/model\_6.e8.raw} \nonumber \\
{\tt spherical\_heat} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/sorted\_xrb.raw} \nonumber
\end{eqnarray}

We use a fortran subroutine
to interpolate the raw data, yielding the model data, $\rho^{\model},
T^{\model}, p^{\model}$, and $X^{\model}$.  The fortran subroutine
then uses an iterative procedure to modify the model data so that it
is thermodynamically consistent with the {\tt MAESTRO} equation of
state (EOS), and also satisfies our chosen hydrostatic equilibrium
(HSE) discretization,
\begin{equation}
\frac{p_r - p_{r-1}}{\Delta r} = \frac{\rho_r + \rho_{r-1}}{2}g_{r-\myhalf},\label{HSE Discretization}
\end{equation}
to a user defined tolerance.  Here are the fortran subroutines for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt test\_convect} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/init\_1d.f90} \nonumber \\
{\tt spherical\_heat} & \rightarrow & {\tt initial\_models/spherical/init\_1d.f90} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/init\_1d.f90} \nonumber
\end{eqnarray}
The model data is not generated at run-time - it must be generated in
advance of running any MAESTRO examples.  The inputs file should point
to the file containing the model data.

>>>>

Here we briefly describe the various routines for generating initial models
for the main \maestro\ problems and how the initial model is used to initialize
both the base state and the full Cartesian state.

@@

We have found that for the best performance, the \maestro\
initialization procedure should be given model data with the same
resolution as the base state resolution, $\Delta r$.  For planar
problems, $\Delta r = \Delta x$.  For multilevel planar problems, we
use $\Delta r$ equal to $\Delta x$ at the finest resolution.  For
spherical problems we set $\Delta r = \Delta x/\mathtt{drdxfac}$.  We
generate our initial model either analytically or from raw data,
$\rho^{\raw}, T^{\raw}, p^{\raw}$, and $X^{\raw}$.  Here is the raw
data file for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt test\_convect} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/model\_6.e8.raw} \nonumber \\
{\tt spherical\_heat} & \rightarrow & \mathrm{none-it~ is~ generated~ analytically} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/sorted\_xrb.raw} \nonumber
\end{eqnarray}

We use a fortran subroutine
to interpolate the raw data, yielding the model data, $\rho^{\model},
T^{\model}, p^{\model}$, and $X^{\model}$.  The fortran subroutine
then uses an iterative procedure to modify the model data so that it
is thermodynamically consistent with the {\tt MAESTRO} equation of
state (EOS), and also satisfies our chosen hydrostatic equilibrium
(HSE) discretization,
\begin{equation}
\frac{p_r - p_{r-1}}{\Delta r} = \frac{\rho_r + \rho_{r-1}}{2}g_{r-\myhalf},\label{HSE Discretization}
\end{equation}
to a user defined tolerance.  Here are the fortran subroutines for each test problem:
\begin{eqnarray}
{\tt test2} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt test\_convect} & \rightarrow & {\tt initial\_models/test2/init\_1d.f90} \nonumber \\
{\tt wdconvect} & \rightarrow & {\tt initial\_models/kepler/init\_1d.f90} \nonumber \\
{\tt spherical\_heat} & \rightarrow & {\tt initial\_models/spherical/init\_1d.f90} \nonumber \\
{\tt xrb} & \rightarrow & {\tt initial\_models/xrb/init\_1d.f90} \nonumber
\end{eqnarray}
The model data is not generated at run-time - it must be generated in
advance of running any \maestro\ examples.  The inputs file should point
to the file containing the model data.

### 2011-06-23 docs/introduction/introduction.tex
use maestrosymbols

\section{History of MAESTRO}

MAESTRO describes the evolution of low Mach number flows that are in
hydrostatic equilibrium.  The idea for MAESTRO grew out of our success
in applying low Mach number combustion methods developed for
terrestrial flames~\cite{DayBell00} to small-scale
astrophysical flows.  Our original small-scale astrophysical
combustion algorithm is detailed in
\begin{itemize}
\item {\em Adaptive Low Mach Number Simulations of Nuclear Flames,}
J. B. Bell, M. S. Day, C. A. Rendleman, S. E. Woosley, \& M. Zingale
2004, JCP, 195, 2, 677 (henceforth BDRWZ)
\end{itemize}

\noindent MAESTRO was developed initially for modeling the convecting
phase in a white dwarf preceding the ignition of a Type Ia supernovae.
As such, we needed to incorporate the compressibility effects due to
large-scale stratification in the star.  The method closest in spirit
to MAESTRO is the pseudo-incompressible method of
Durran~\cite{durran}, developed for terrestrial atmospheric flows
(assuming an ideal gas).  Part of the complexity of the equations in
MAESTRO stems from the need to descibe a general equation of state.
Additionally, since reactions can significantly alter the hydrostatic
structure of a star, we incorporated extensions that capture the
expansion of the background state~\cite{almgren:2000}.  The low Mach
number equations for stellar flows were developed in a series of
papers leading up to the first application to this problem:
\begin{itemize}
\item {\em Low Mach Number Modeling of Type Ia
  Supernovae. I. Hydrodynamics,} A. S. Almgren, J. B. Bell, 
  C. A. Rendleman, \& M. Zingale 2006, ApJ, 637, 922 (henceforth
  paper I)
\item {\em Low Mach Number Modeling of Type Ia Supernovae. II. Energy
  Evolution,} A. S. Almgren, J. B. Bell, C. A. Rendleman, \& M. Zingale
  2006, ApJ, 649, 927 (henceforth paper II)
\item {\em Low Mach Number Modeling of Type Ia Supernovae. III. Reactions,}
  A. S. Almgren, J. B. Bell, A. Nonaka, \& M. Zingale
  2008, ApJ, 684, 449 (henceforth paper III)
\item {\em Low Mach Number Modeling of Type Ia Supernovae. IV. White Dwarf Convection,}
  M. Zingale, A. S. Almgren, J. B. Bell, A. Nonaka, \& S. E. Woosley
  2009, ApJ, 704, 196 (henceforth paper IV)
\end{itemize}

@@

Extensions to variable-density incompressible
flows~\cite{bellMarcus:1992b} involve a slightly different
decomposition of the velocity field and, as a result, a slightly
different Poisson equation.  There is also a variety of different ways
to express what is being projected~\cite{almgren:bell:crutchfield},
and different discretizations of the divergence and gradient operators
lead to slightly different mathematical properties of the methods
(leading to ``approximate
projections''~\cite{almgrenBellSzymczak:1996}).  Finally, for
second-order methods, two projections are typically done per timestep.
The first (the `MAC' projection~\cite{bellColellaHowell:1991})
operates on the half-time, edge-centered advective velocities, making
sure that they satisfy the divergence constraint.  These advective
velocities are used to construct the fluxes through the interfaces to
advance the solution to the new time.  The second/final projection
operates on the cell-centered velocities at the new time, again
enforcing the divergence constraint.  The MAESTRO algorithm performs
both of these projections.

The MAESTRO algorithm builds upon these ideas, using a different
velocity constraint equation that captures the compressibility
due to local sources and large-scale stratification.

@@

Throughout the papers describing MAESTRO, we've largely kept our
notation consistent.  Table~\ref{table:sym} defines the
frequently-used quantities and provides their units.

>>>>

\section{History of \maestro}

\maestro\ describes the evolution of low Mach number flows that are in
hydrostatic equilibrium.  The idea for \maestro\ grew out of our success
in applying low Mach number combustion methods developed for
terrestrial flames~\cite{DayBell00} to small-scale
astrophysical flows.  Our original small-scale astrophysical
combustion algorithm is detailed in
\begin{itemize}
\item {\em Adaptive Low Mach Number Simulations of Nuclear Flames,}
J. B. Bell, M. S. Day, C. A. Rendleman, S. E. Woosley, \& M. Zingale
2004, JCP, 195, 2, 677 (henceforth BDRWZ)
\end{itemize}

\noindent \maestro\ was developed initially for modeling the convecting
phase in a white dwarf preceding the ignition of a Type Ia supernovae.
As such, we needed to incorporate the compressibility effects due to
large-scale stratification in the star.  The method closest in spirit
to \maestro\ is the pseudo-incompressible method of
Durran~\cite{durran}, developed for terrestrial atmospheric flows
(assuming an ideal gas).  Part of the complexity of the equations in
\maestro\ stems from the need to descibe a general equation of state.
Additionally, since reactions can significantly alter the hydrostatic
structure of a star, we incorporated extensions that capture the
expansion of the background state~\cite{almgren:2000}.  The low Mach
number equations for stellar flows were developed in a series of
papers leading up to the first application to this problem:
\begin{itemize}
\item {\em Low Mach Number Modeling of Type Ia
  Supernovae. I. Hydrodynamics,} A. S. Almgren, J. B. Bell, 
  C. A. Rendleman, \& M. Zingale 2006, ApJ, 637, 922 (henceforth
  paper I)
\item {\em Low Mach Number Modeling of Type Ia Supernovae. II. Energy
  Evolution,} A. S. Almgren, J. B. Bell, C. A. Rendleman, \& M. Zingale
  2006, ApJ, 649, 927 (henceforth paper II)
\item {\em Low Mach Number Modeling of Type Ia Supernovae. III. Reactions,}
  A. S. Almgren, J. B. Bell, A. Nonaka, \& M. Zingale
  2008, ApJ, 684, 449 (henceforth paper III)
\item {\em Low Mach Number Modeling of Type Ia Supernovae. IV. White Dwarf Convection,}
  M. Zingale, A. S. Almgren, J. B. Bell, A. Nonaka, \& S. E. Woosley
  2009, ApJ, 704, 196 (henceforth paper IV)
\end{itemize}

@@

Extensions to variable-density incompressible
flows~\cite{bellMarcus:1992b} involve a slightly different
decomposition of the velocity field and, as a result, a slightly
different Poisson equation.  There is also a variety of different ways
to express what is being projected~\cite{almgren:bell:crutchfield},
and different discretizations of the divergence and gradient operators
lead to slightly different mathematical properties of the methods
(leading to ``approximate
projections''~\cite{almgrenBellSzymczak:1996}).  Finally, for
second-order methods, two projections are typically done per timestep.
The first (the `MAC' projection~\cite{bellColellaHowell:1991})
operates on the half-time, edge-centered advective velocities, making
sure that they satisfy the divergence constraint.  These advective
velocities are used to construct the fluxes through the interfaces to
advance the solution to the new time.  The second/final projection
operates on the cell-centered velocities at the new time, again
enforcing the divergence constraint.  The \maestro\ algorithm performs
both of these projections.

The \maestro\ algorithm builds upon these ideas, using a different
velocity constraint equation that captures the compressibility
due to local sources and large-scale stratification.

@@

Throughout the papers describing \maestro, we've largely kept our
notation consistent.  Table~\ref{table:sym} defines the
frequently-used quantities and provides their units.

### 2011-06-23 docs/lo_density/lo_density.tex
caption fix
use maestrosymbols

These are working notes for the low density parameters in {\tt MAESTRO}.  In low density
regions, we modify the behavior of the algorithm.  Here is a summary of some parameters,
and a brief description of what they do.
\begin{itemize}
\item {\tt base\_cutoff\_density}, $\rho_{\rm base}$, (real):\\ \\
Essentially controls the lowest density allowed in the simulation and modifies the behavior
of several modules.
\item {\tt base\_cutoff\_density\_coord(:)} (integer array):\\ \\
For each level in the radial base state array, this is the coordinate of the first cell
where $\rho_0 < \rho_{\rm base}$.  Slightly more complicated for multilevel problems.
\item {\tt anelastic\_cutoff}, $\rho_{\rm anelastic}$, (real):\\ \\
If $\rho_0 < \rho_{\rm anelastic}$, we modify the computation of $\beta_0$ in the
divergence constraint.
\item {\tt anelastic\_cutoff\_coord(:)} (integer array):\\ \\
Anelastic cutoff analogy of {\tt base\_cutoff\_density\_coord(:)}.
\item {\tt burning\_cutoff\_density}, $\rho_{\rm burning}$,  (real):\\ \\
If $\rho < \rho_{\rm burning}$, don't call the burner in this cell.
\item {\tt burning\_cutoff\_density\_coord(:)} (integer array):\\ \\
Burning cutoff analogy of {\tt base\_cutoff\_density\_coord(:)}.
\item {\tt buoyancy\_cutoff\_factor} (real):\\ \\
When computing velocity forcing, set the buoyance term ($\rho-\rho_0$) to 0 if 
$\rho < $ {\tt buoyancy\_cutoff\_factor} * {\tt base\_cutoff\_density}.
\item {\tt do\_eos\_h\_above\_cutoff} (logical):\\ \\
If {\rm true}, at the end of the advection step, for each cell where 
$\rho < \rho_{\rm base}$, recompute $h = h(\rho,p_0,X)$.
\end{itemize}

@@

\subsection{Single-Level Planar or any Spherical}
Here the base state exists as a single one-dimensional array with constant grid
spacing $\Delta r$.  Basically, we set the corresponding coordinate equal to $r$ as soon 
as $\rho_0(r)$ is less than  or equal to that particular cutoff value.
See Figure \ref{Fig:Cutoff} for a graphical representation.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[hpb]
\centering
\includegraphics[width=4.0in]{\lodensfigpath/cutoff}\hspace{0.2in}
\caption{Image of how the cutoff density and cutoff coordinates
are related for single-level planar and all spherical problems.}
\label{Fig:Cutoff}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

These are working notes for the low density parameters in \maestro.
In low density regions, we modify the behavior of the algorithm.  Here
is a summary of some parameters, and a brief description of what they
do.
\begin{itemize}
\item {\tt base\_cutoff\_density}, $\rho_{\rm base}$, (real):\\ \\
Essentially controls the lowest density allowed in the simulation and modifies the behavior
of several modules.
\item {\tt base\_cutoff\_density\_coord(:)} (integer array):\\ \\
For each level in the radial base state array, this is the coordinate of the first cell
where $\rho_0 < \rho_{\rm base}$.  Slightly more complicated for multilevel problems.
\item {\tt anelastic\_cutoff}, $\rho_{\rm anelastic}$, (real):\\ \\
If $\rho_0 < \rho_{\rm anelastic}$, we modify the computation of $\beta_0$ in the
divergence constraint.
\item {\tt anelastic\_cutoff\_coord(:)} (integer array):\\ \\
Anelastic cutoff analogy of {\tt base\_cutoff\_density\_coord(:)}.
\item {\tt burning\_cutoff\_density}, $\rho_{\rm burning}$,  (real):\\ \\
If $\rho < \rho_{\rm burning}$, don't call the burner in this cell.
\item {\tt burning\_cutoff\_density\_coord(:)} (integer array):\\ \\
Burning cutoff analogy of {\tt base\_cutoff\_density\_coord(:)}.
\item {\tt buoyancy\_cutoff\_factor} (real):\\ \\
When computing velocity forcing, set the buoyance term ($\rho-\rho_0$) to 0 if 
$\rho < $ {\tt buoyancy\_cutoff\_factor} * {\tt base\_cutoff\_density}.
\item {\tt do\_eos\_h\_above\_cutoff} (logical):\\ \\
If {\rm true}, at the end of the advection step, for each cell where 
$\rho < \rho_{\rm base}$, recompute $h = h(\rho,p_0,X)$.
\end{itemize}

@@

\subsection{Single-Level Planar or any Spherical}
Here the base state exists as a single one-dimensional array with constant grid
spacing $\Delta r$.  Basically, we set the corresponding coordinate equal to $r$ as soon 
as $\rho_0(r)$ is less than  or equal to that particular cutoff value.
See Figure \ref{Fig:Cutoff} for a graphical representation.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[hpb]
\centering
\includegraphics[width=4.0in]{\lodensfigpath/cutoff}\hspace{0.2in}
\caption[Cutoff density and coordinates]{Image of how the cutoff density and cutoff coordinates
are related for single-level planar and all spherical problems.}
\label{Fig:Cutoff}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
